draft

### 2025-05-19 draft.tex
docs: finish introduction revamp

% Created 2025-05-19 Mon 17:52
% Intended LaTeX compiler: pdflatex
\documentclass{article}

@@

A CellProfiler pipeline struggles integrating well with other computational tools. Building plug-ins to add functionality is time and effort-consuming challenges that require an understanding of its interfaces. Treating it as a black box and using its output precludes the addition of custom-made image filters. Lastly, CellProfiler depends on many number of packages from both Python and Java, amplifying the risk of dependency conflicts. Containers mitigate this problem, but not without increased complexity and caveats. To summarize, CellProfiler is an all-or-nothing tool, integrating other tools demands more work and time than most scientists can spare.

Object featurization is not exclusive of CellProfiler. For example, scikit-image can produce some overlapping features. Tools like ScaleFex \citep{comoletHighlyEfficientScalable2024,einarolafssonSpaCr2025} generate features, but they have limitations: First, their implementation is completely independent and thus cannot reproduce existing datasets, potentially impacting the biological interpretations. Moreover, it enforces a specific data or metadata structure, which also adds complexity to the setup. Lastly, a common pattern profilers akin to ScaleFex is that they enforce a specific shape for the input data and require wrangling with the data before processing a single image. Some alternatives, such as ScaleFex, are designed for cloud usage, neglectling a significant number of cases where local compute suffices. We thus believe that there is the need for a tool aimed at generalist and high throughput problems not being solved by the current software ecosystem.
\subsection{Engineered features in the era of deep learning}
\label{sec:org9dc3dfa}
The unprecedented evolution of deep learning methods in recent years has made available many different ways to process imaging data. Nevertheless, these are limited in their capacity to provide biological interpretation \citep{moenDeepLearningCellular2019}. Engineered features can also outperform deep learning ones under certain conditions \citep{tangMorphologicalProfilingDrug2024}. Even as deep learning features improve over time, they still require complex interpretability methods, in the rare case where they can be interpreted. Deep Learning also has a higher barrier of entry, since most models require annotating data for training and testing. The foundational models whose goal is to bypass this step do not generalize well for out-of-distribution data. Unlike deep learning-based features, whose interpretation requires complex methods such as counterfactuals or attention maps, most of the engineered features have a mathematical definition that is vastly simpler to translate into biological insights.

The core target users of such a tool would be biologists and data scientists who need reproducibility and data within a few lines of code. As a whole, there is still a role to be played for engineered features to, at the very least, complement deep learning ones. The community is missing a generalist tool that converts images and masks into representative vectors without the need of interfaces or cloud compute and that integrates seamlessly with other scientific libraries. This is the role that we envision cp\_measure can occupy in the scientific software ecosystem.
\section{Extracting interpretable features}
\label{sec:org61842b5}
In cp\_measure, we isolated the CellProfiler measurements and categorize them based on their inputs: One object and one imaging channel (e.g., intensity, texture), one object and multiple channels (such as correlation features) and multiple objects (e.g., number of neighbors).

>>>>

% Created 2025-05-19 Mon 18:10
% Intended LaTeX compiler: pdflatex
\documentclass{article}

@@

Integrating a CellProfiler pipelines with other tools is a struggle. Building plug-ins to add functionality is time and effort-consuming challenges that require an understanding of its interfaces. Treating it as a black box and using its output precludes the addition of custom-made image filters. Lastly, CellProfiler depends on many number of packages from both Python and Java, amplifying the risk of dependency conflicts. Containers mitigate this problem, but not without increased complexity and caveats. To summarize, CellProfiler is an all-or-nothing tool, integrating other tools demands more work and time than most scientists can spare.

Object featurization is not exclusive of CellProfiler. For example, scikit-image can produce some overlapping features. Tools like ScaleFex \citep{comoletHighlyEfficientScalable2024,einarolafssonSpaCr2025} generate features, but they have limitations: First, their implementation is completely independent and thus cannot reproduce existing datasets, potentially impacting the biological interpretations. Moreover, a common portability problem is that these tools enforce a specific shape for the input data and require wrangling with the data before processing a single image. Some alternatives, such as ScaleFex, are designed for cloud usage first, and may neglect a significant number of cases where local compute suffices. We thus believe that there is the need for a tool aimed at generalist and high throughput problems not being solved by the current software ecosystem.
\subsection{Engineered features in the era of deep learning}
\label{sec:org9dc3dfa}
The unprecedented evolution of deep learning methods in recent years has made available many different ways to process imaging data. Nevertheless, these are limited in their capacity to provide biological interpretation \citep{moenDeepLearningCellular2019}. Engineered features can also outperform deep learning ones under certain conditions \citep{tangMorphologicalProfilingDrug2024}. Even as deep learning features improve over time, they still require complex interpretability methods, even if they manage to be interpretable at all. Compared to engineered features, deep learning approaches have a higher barrier of entry, as most models require data annotating. Foundational models, whose goal is to lower this barrier, fail to generalize well for out-of-distribution data. Unlike deep learning-based features, whose interpretation requires complex methods like counterfactuals or attention maps, most engineered features have a mathematical definition that is faster to calculate and simpler to translate into biological concepts.

The intended audience for such a featurization tool would be, in addition to biologists, data scientists striving to achieve reproducible data analyses within a few lines of code. As a whole, engineered features still have the role of, at the minimum, complementing deep learning ones. The bioimage software ecosystems lacks a modular tool to featurize images into interpretable features with no compulsory graphical interfaces or cloud compute. Ideally it should also harmonize with other scientific libraries. With all this in mind we have developed cp\_measure.
\section{Extracting interpretable features}
\label{sec:org61842b5}
In cp\_measure, we isolated the CellProfiler measurements and categorize them based on their inputs: One object and one imaging channel (e.g., intensity, texture), one object and multiple channels (such as correlation features) and multiple objects (e.g., number of neighbors).
